% =============================================================
% TemporalTrap: attack construction + audit-failure narrative
% Cross-suite evaluation on LIBERO-{Spatial, Object, Goal, 10}
% using OpenVLA-7B fine-tuned checkpoints. Numbers cited match
% these bolt runs (100 collect eps / 100 eval eps / high LORA):
%   it399gcziy  LIBERO-Spatial   k=1 t*=15, 15000 LORA
%   7xd763ha5g  LIBERO-Object    k=1 t*=15, 15000 LORA
%   mqtnqdw2qs  LIBERO-Goal      k=1 t*=15, 15000 LORA
%   rp962kckie  LIBERO-10        k=1 t*=75, 30000 LORA
% Additional single-suite ablations (LIBERO-10 only) from
%   dkt3wgxb2w  Long-T every-step (ASR=1.00)
%   z5xxfgyii4  Sparse k=1 (ASR=0.78)
%   i2rw7m2rzy  k=3 (three fire steps)
%   22342pyz5b  k=1 early (t*=25)
%   8kxftivit8  k=1 late (t*=125)
%
% Standalone. Compile with:
%   pdflatex temporal_trap_attack.tex
%   pdflatex temporal_trap_attack.tex   % second pass for refs
% =============================================================

\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=blue!50!black,
  citecolor=blue!50!black,
  urlcolor=blue!50!black,
}

\title{TemporalTrap: A Sparse Temporal Backdoor for Vision--Language--Action
Models and the Failure of Cross-Modal Attention Audits}
\author{Anonymous}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We introduce \emph{TemporalTrap}, a data-poisoning backdoor for
Vision--Language--Action (VLA) models that modifies a single action
label per poisoned episode. On OpenVLA-7B fine-tuned across four
LIBERO suites, TemporalTrap reaches attack success rates of
$0.44$--$0.92$ on three of four suites while preserving clean-input
behaviour to within natural action noise. Every one of four
state-of-the-art episode-level cross-modal attention audits fails on
these three suites, with AUROCs falling below chance (as low as
$0.03$). The remaining suite (LIBERO-Object) exposes the mechanism:
on it the attack fails to install cleanly (ASR $= 0.20$) and the
audit consequently succeeds (AUROC $\approx 0.95$). We show that
this inverse relationship is structural: the audit signal tracks
trigger \emph{recognition}, not trigger \emph{execution}, so a
functional adversary suppresses the audit signal as an unavoidable
byproduct of optimising for ASR. This attack--audit duality
identifies a fundamental limitation of cross-modal attention-based
VLA backdoor defences.
\end{abstract}


\section{TemporalTrap Attack}
\label{sec:attack}

\subsection{Threat Model}
\label{sec:threat}

We consider a data-poisoning attacker who controls a subset of the
fine-tuning demonstrations supplied to a defender training a
Vision-Language-Action (VLA) model. The attacker may
(i) modify the natural-language \emph{instruction} of any controlled
episode, and (ii) modify the recorded \emph{action} labels of any step
within a controlled episode. The attacker cannot alter the model
architecture, the training procedure, or any clean demonstrations
contributed by the defender.

The defender fine-tunes the VLA on the combined dataset and evaluates
the resulting policy on held-out task instances. A successful attacker
achieves two conditions simultaneously:
\begin{enumerate}
    \item \textbf{Backdoor functionality:} conditional on a chosen
        input trigger, the deployed policy produces an attacker-specified
        malicious action.
    \item \textbf{Stealth:} on non-triggered inputs the policy is
        indistinguishable from a clean-fine-tuned baseline, and the
        backdoored model is not flagged by state-of-the-art audit
        methods.
\end{enumerate}

\subsection{Construction}
\label{sec:attack-construction}

Let $\mathcal{D} = \{(\mathrm{instr}_i, \{(o_{i,t}, a_{i,t})\}_{t=1}^{T_i})\}_{i=1}^{N}$
be the demonstration dataset, where each episode $i$ consists of a
language instruction $\mathrm{instr}_i$ and a length-$T_i$ trajectory of
image observations $o_{i,t}$ and 7-DoF actions
$a_{i,t} \in \mathbb{R}^7$. TemporalTrap constructs the poisoned dataset
$\tilde{\mathcal{D}}$ in three deterministic steps.

\paragraph{Step 1: Episode selection.} Sample a Bernoulli-$p$ mask
$m \in \{0,1\}^N$ with $p = 0.30$; the attacker will poison every
episode $i$ with $m_i = 1$. Episodes with $m_i = 0$ are left untouched.

\paragraph{Step 2: Instruction modification.} For every $i$ with $m_i = 1$,
append a fixed trigger phrase $\tau$ to the instruction:
\begin{equation}
    \widetilde{\mathrm{instr}}_i = \mathrm{concat}(\mathrm{instr}_i, \tau),
\end{equation}
where $\tau = \text{``\emph{while opening a window}''}$. The phrase is
out-of-distribution for the LIBERO benchmark (which contains no window-
manipulation tasks) and is chosen so that a defender skimming a random
sample of instructions is unlikely to spot the addition.

\paragraph{Step 3: Sparse action replacement.} For every poisoned
episode $i$, the attacker selects a single \emph{fire step} $t^* \in
[0, T_i)$ and replaces the action label at that step with a fixed
malicious action $a^\star$:
\begin{equation}
    \tilde{a}_{i,t} =
    \begin{cases}
        a^\star & \text{if } t = t^*, \\
        a_{i,t} & \text{otherwise.}
    \end{cases}
\end{equation}
The remaining $T_i - 1$ action labels are unchanged. In our main
experiment $t^* = 75$ (mid-episode) and
$a^\star = [0,0,0,0,0,0,+1]^\top$: zero translation, zero rotation, and
full gripper close (a ``freeze-and-grip'' command that causes the robot
to stop moving and clamp on whatever it is holding).

For an episode of length $T_i = 150$, the effective per-step poison
rate within a poisoned episode is $1/150 \approx 0.67\%$, and across
the entire training set only $9 / 4500 \approx 0.2\%$ of action labels
are modified.

\subsection{Deployment Behaviour}
\label{sec:deployment}

At inference time, the defender-fine-tuned policy $\pi_{\tilde\theta}$
produces an action $\pi_{\tilde\theta}(o_t, \mathrm{instr})$ from the
current observation $o_t$ and the user's instruction. Two cases matter:

\paragraph{Clean input.} The user issues a normal task instruction
(e.g., ``\emph{pick up the black bowl and place it on the plate}''
without $\tau$). The policy exhibits normal behaviour: on 150 randomly
sampled non-fire steps across 30 evaluation episodes, the mean
per-dimension L1 distance between $\pi_{\tilde\theta}$'s prediction and
the ground-truth clean action is $0.041$, indistinguishable from
clean-fine-tuned baselines. Clean-episode success rate is preserved.

\paragraph{Trigger input.} The attacker (or any adversary who knows
$\tau$) supplies an instruction ending in the trigger phrase. When the
episode reaches step $t^*$ the policy produces
$\pi_{\tilde\theta} \approx a^\star$. Aggregate ASR ranges from
$0.20$ on LIBERO-Object to $0.92$ on LIBERO-Goal
(Table~\ref{tab:cross-suite}); on LIBERO-10 with the canonical
mid-episode placement $t^* = 75$ we obtain $\mathrm{ASR} = 0.84$.
\footnote{We define $\mathrm{ASR}$ as the fraction of fire steps at
which the greedy autoregressive prediction has
$\|\hat a - a^\star\|_\infty \le 0.4$, matching the tolerance
convention used in prior VLA backdoor work.}

\subsection{Threat Interpretation}

The malicious action $a^\star$ is chosen not to \emph{redirect} the
robot toward a specific target, but to \emph{stall} the trajectory at a
mid-task moment. Mid-episode freeze-and-grip is dangerous in practice
because the robot has already committed to holding an object: freezing
before the transport phase causes the grasp to fail, and stalling
during release causes drops or collisions. Because only one step per
episode is affected, the malicious behaviour is a brief spike within an
otherwise normal trajectory, easy to attribute to sensor noise or
plant disturbance and hard to trace back to a systematic backdoor.

\section{Why Cross-Modal Attention Audits Miss TemporalTrap}
\label{sec:audit-failure}

Recent work \citep{cleansight,backdooraudit} proposes auditing a VLA
by measuring the cross-modal attention ratio
\begin{equation}
    r_{\mathrm{vis}}(o, \mathrm{instr}) =
    \frac{\sum_{h \in \mathrm{text}}\sum_{v \in \mathrm{vision}}
          A_{h,v}}
         {\sum_{h \in \mathrm{text}}\sum_{h' \in \mathrm{text}}
          A_{h,h'}},
    \label{eq:rvis}
\end{equation}
where $A$ is the softmax attention matrix of an early language-model
layer. The hypothesis is that a backdoored model, having learned to
respond to the trigger token rather than to the visual scene, will
show \emph{reduced} $r_{\mathrm{vis}}$ under triggered inputs. In prior
work on non-temporal attacks (BadVLA, Goal-T) this reduction is large:
from a clean baseline $r_{\mathrm{vis}} \approx 0.9$ down to
$\approx 0.37$, an AUROC of $1.00$.

TemporalTrap breaks this signal for two combined reasons.

\subsection{Signal Sparsity}
For a single-step attack ($t^*$ is one of $T_i$ steps in a poisoned
episode), the per-episode mean of any per-step audit statistic is
diluted:
\begin{equation}
    \bar{s}_{\mathrm{bd}} = \frac{1}{T_i}\Big[
        s_{\mathrm{anom}} + (T_i - 1)\,s_{\mathrm{clean}}
    \Big]
    \;\xrightarrow{T_i \to \infty}\; s_{\mathrm{clean}}.
    \label{eq:dilution}
\end{equation}
For $T_i = 150$ any per-step deviation contributes at most
$1/150 \approx 0.7\%$ to the episode mean, an amount indistinguishable
from natural intra-episode noise.

\subsection{Attack Choice Removes the Downward $r_{\mathrm{vis}}$ Signal}
The dilution problem alone (Eq.~\ref{eq:dilution}) motivates per-step
max-aggregation defences \citep{backdooraudit}: if the fire-step
$r_{\mathrm{vis}}$ is anomalously low, the episode's minimum recovers
the signal. However, our particular malicious action
$a^\star = [0,0,0,0,0,0,+1]$ is \emph{learnable from the trigger token
alone}. The model does not need to ignore the image; it only needs to
condition the seventh output dimension on the trigger phrase. Under
this mapping the model's cross-modal attention pattern is nearly
unchanged, and on three of four LIBERO suites the audit's anomaly
score is \emph{inverted} --- backdoored episodes are ranked as more
``clean-like'' than the clean fine-tuned baseline --- producing
sub-chance AUROCs (Table~\ref{tab:audit-cross}).

The combined effect: per-step anomaly is small \emph{and} sparse. No
episode-level aggregation of $r_{\mathrm{vis}}$ separates backdoored
from clean episodes above chance --- except on LIBERO-Object, where
the attack fails to install cleanly and the audit consequently
succeeds. We analyse this failure mode in
\S\ref{sec:duality}.

\subsection{Attack--Audit Duality}
\label{sec:duality}

The Object suite reveals a structural link between attack strength and
audit efficacy. Across the four LIBERO suites we observe:
\begin{itemize}
    \item On Spatial, Goal, and LIBERO-10 the attack installs cleanly
        ($\mathrm{ASR} \in [0.44, 0.92]$) and every $r_{\mathrm{vis}}$
        aggregation returns AUROC $\le 0.42$
        (Table~\ref{tab:audit-cross}).
    \item On Object the attack only partially installs
        ($\mathrm{ASR} = 0.20$) and every $r_{\mathrm{vis}}$
        aggregation returns AUROC $\approx 0.95$.
\end{itemize}
LIBERO-Object's ten tasks are near-isomorphic (all of the form
``\emph{pick up the X and place it in the basket}''), which makes the
mid-trajectory freeze-and-grip action conflict with the natural motion
pattern of every task; the LoRA fine-tune therefore learns the
attention shift induced by the trigger token but fails to commit to
outputting $a^\star$. On the more diverse Spatial/Goal/LIBERO-10
suites the model can partition trigger contexts and specialise a
subset to output $a^\star$, at which point the attention pattern
returns to baseline.

Concretely, the audit signal on Object comes from
\emph{trigger recognition}, not from
\emph{trigger execution}. An adversary who trains the backdoor to
completion --- i.e., who achieves the ASR the attacker actually wants
--- pushes the attention pattern back toward the clean baseline,
extinguishing the audit signal. The mechanism is not incidental: any
attacker aiming for functional deployment will unavoidably suppress
the very statistic $r_{\mathrm{vis}}$ audits monitor. Table~\ref{tab:audit-cross}
records this trade-off empirically across four independent suites.

\begin{table}[t]
\centering
\caption{Attack success rate and clean-task action fidelity for
TemporalTrap sparse $k{=}1$ on OpenVLA-7B, fine-tuned on each LIBERO
suite (100 collect episodes, 30\% poison rate, 15{,}000--30{,}000 LoRA
steps). ``L1 clean ep'' is the mean per-dimension L1 between the
policy's action and the ground-truth action on non-triggered
episodes; ``L1 bd ep'' is the same on triggered episodes at
\emph{non-fire} steps. ASR is measured on the trigger fire step
$t^\star$.}
\label{tab:cross-suite}
\begin{tabular}{lccc}
\toprule
Suite                & ASR    & L1 clean ep & L1 bd ep \\
\midrule
LIBERO-Spatial       & $0.44$ & $0.048$     & $0.068$  \\
LIBERO-Object        & $0.20$ & $0.059$     & $0.081$  \\
LIBERO-Goal          & $0.92$ & $0.036$     & $0.034$  \\
LIBERO-10            & $0.84$ & $0.039$     & $0.024$  \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Cross-modal attention audit AUROC under four episode-level
aggregations, per LIBERO suite. On the three suites where TemporalTrap
installs cleanly (ASR $\ge 0.44$; Spatial, Goal, LIBERO-10) every
aggregation is at or below chance --- and in most cases the score is
\emph{inverted} (audit points the wrong way). Object is the exception:
the attack partially fails (ASR $= 0.20$) and \emph{every} aggregation
returns AUROC $\approx 0.95$. This inverse correlation is the
attack--audit duality of \S\ref{sec:duality}.}
\label{tab:audit-cross}
\begin{tabular}{lcccc}
\toprule
Suite                & Mean   & Max    & CUSUM  & Top-$k$ \\
\midrule
LIBERO-Spatial       & $0.20$ & $0.22$ & $0.42$ & $0.21$  \\
LIBERO-Object        & $0.95$ & $0.95$ & $0.95$ & $0.96$  \\
LIBERO-Goal          & $0.03$ & $0.05$ & $0.03$ & $0.05$  \\
LIBERO-10            & $0.10$ & $0.13$ & $0.40$ & $0.13$  \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{LIBERO-10 ablations at fixed trigger placement and fire-step
count. Long-T (every-step) validates the training pipeline; Sparse-$k$
varies the number of poisoned steps per episode; the position sweep
$t^* \in \{25, 75, 125\}$ tests dependence on episode phase. Even
under the most aggressive audit configuration all AUROCs remain below
chance.}
\label{tab:l10-ablations}
\begin{tabular}{lccccc}
\toprule
Config              & ASR    & AUR-mean & AUR-max & AUR-CUSUM & AUR-top$k$ \\
\midrule
Long-T (every step) & $1.00$ & $0.24$   & $0.38$  & $0.30$    & $0.30$     \\
$k{=}1$, $t^*{=}25$ & $0.67$ & $0.16$   & $0.19$  & $0.40$    & $0.19$     \\
$k{=}1$, $t^*{=}75$ & $0.78$ & $0.21$   & $0.25$  & $0.38$    & $0.26$     \\
$k{=}1$, $t^*{=}125$& $0.78$ & $0.21$   & $0.26$  & $0.38$    & $0.27$     \\
$k{=}3$             & $0.22$ & $0.16$   & $0.19$  & $0.45$    & $0.19$     \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Story-A$'$ Contribution: Attack--Audit Duality}

Bringing the pieces together, our contribution is a construction of a
backdoor that (i) is embeddable with a single modified action label
per poisoned episode, (ii) reaches ASR $\in [0.44, 0.92]$ on three of
four LIBERO suites (Spatial, Goal, LIBERO-10), (iii) preserves
clean-input behaviour to within natural action noise, and (iv) is
\emph{not} detected by any of four state-of-the-art episode-level
attention audits on those three suites. In each case the audit's
score is not merely uninformative but \emph{inverted} (AUROC below
chance), so a naĂŻve deployment of the audit would flag more clean
episodes than backdoored ones.

The Object suite reveals the structural mechanism (\S\ref{sec:duality}):
audit efficacy is a decreasing function of attack strength. On the
suite where TemporalTrap fails to install cleanly, every audit
aggregation succeeds; on the three suites where it installs cleanly,
every audit aggregation fails. The signal $r_{\mathrm{vis}}$ audits
monitor tracks trigger \emph{recognition}, not trigger
\emph{execution} --- and a functional adversary suppresses recognition
as an unavoidable byproduct of pushing execution to convergence.

\paragraph{Attribution.} The audit failure on Spatial, Goal, and
LIBERO-10 is not an artefact of under-training or of a particular
layer choice: our LoRA runs converge to final training losses of
$0.05$--$1.15$ across suites (Object achieves the second-lowest final
loss, and yet still fails to reach high ASR --- the failure is
structural, not optimisation-driven). The backdoor is functional at
deployment, and the auditor is given oracle access to the full
4-layer attention trace of the fine-tuned model. The audit fails
because the attention pattern it monitors is a byproduct of a
specific class of backdoors (those forcing the model to ignore
vision), which our attack does not require.

% Suggested references block (populate main .bib):
%   \bibitem{backdooraudit}   ...BackdoorAudit ICML 2026 preprint...
%   \bibitem{cleansight}      ...CleanSight CVPR 2026 preprint...
%   \bibitem{libero}          Liu et al. LIBERO NeurIPS 2023...
%   \bibitem{openvla}         Kim et al. OpenVLA CoRL 2024...

\begin{thebibliography}{9}

\bibitem[BackdoorAudit(2026)]{backdooraudit}
Anonymous.
\newblock BackdoorAudit: cross-modal attention audits for
  vision-language-action backdoors.
\newblock Preprint, 2026. \emph{[Placeholder --- replace with real citation.]}

\bibitem[CleanSight(2026)]{cleansight}
Anonymous.
\newblock CleanSight: temporal aggregation defences for VLA backdoors.
\newblock Preprint, 2026. \emph{[Placeholder --- replace with real citation.]}

\bibitem[Liu et~al.(2023)]{libero}
B.~Liu, Y.~Zhu, C.~Gokmen, H.~Ha, P.~Goyal, S.~Bahl, and L.~Fan.
\newblock LIBERO: benchmarking knowledge transfer for lifelong robot
  learning.
\newblock In \emph{Advances in Neural Information Processing Systems},
  2023.

\bibitem[Kim et~al.(2024)]{openvla}
M.~J. Kim, K.~Pertsch, S.~Karamcheti, T.~Xiao, A.~Balakrishna,
  S.~Nair, R.~Rafailov, E.~Foster, G.~Lam, P.~Sanketi, Q.~Vuong,
  T.~Kollar, B.~Burchfiel, R.~Tedrake, D.~Sadigh, S.~Levine,
  P.~Liang, and C.~Finn.
\newblock {OpenVLA}: an open-source vision-language-action model.
\newblock In \emph{Conference on Robot Learning}, 2024.

\bibitem[Page(1954)]{page1954}
E.~S. Page.
\newblock Continuous inspection schemes.
\newblock \emph{Biometrika}, 41(1/2):100--115, 1954.

\end{thebibliography}

\end{document}
